---
title: "earthquakeMap_Functions"
author: "Cody Schulz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
fig_width: 8
fig_height: 6
vignette: >
  %\VignetteIndexEntry{earthquakeMap_Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# earthquakeMap

earthquakeMap helps users visualize over 4,000 years of earthquake statistics in both time and space.

## Installation

earthquakeMap can be installed using `devtools`:
```{r devtools, eval = FALSE}
library(devtools)
install_github("codyschulz/earthquakeMap")
library(earthquakeMap)
```
## Load Packages

```{r load_packages, results = 'hide', warning = FALSE, message = FALSE}
## Open packages
libs <- c("readr", "magrittr", "dplyr", "earthquakeMap", "ggplot2", "lubridate")
lapply(libs, require, character.only = TRUE)
knitr::opts_chunk$set(fig.width=8, fig.height=6) 
```

## `earthquake_read()`

`earthquake_read()` reads the raw earthquake data in as a `dplyr` `tbl_df`.
```{r earthquake_read}

## Open dataset
x <- earthquake_read()
```

## `eq_location_clean()`

`eq_location_clean()` cleans the LOCATION_NAME variable for use in plotting. The country name and extraneous location information are removed and the string is put in proper case.

```{r eq_location_clean}
head(x$LOCATION_NAME)
y <- eq_location_clean(x)
head(y$LOCATION_NAME)
```

## `eq_clean_data()`

`eq_clean_data()` readies the raw earthquake data for plotting. It subsets the data to only include earthquakes happening in AD years; converts the EQ_PRIMARY, TOTAL_DEATHS, LATITUDE, LONGITUDE to numeric; creates a date variable; and runs `eq_location_clean()`.
```{r eq_clean_data}
str(x[,c("EQ_PRIMARY", "TOTAL_DEATHS", "LATITUDE", "LONGITUDE")])
range(x$YEAR)
y1 <- eq_clean_data(x)
str(y1[,c("EQ_PRIMARY", "TOTAL_DEATHS", "LATITUDE", "LONGITUDE")])
range(y1$YEAR)
head(y1$date)
```

## `geom_timeline()`

`geom_timeline()` function displays earthquakes on a timeline, with point size corresponding to magnitude and point color to the number of fatalities. Each country selected is given its own timeline. Below is a plot of all earthquakes in the US and Mexico since 2005.


```{r geom_timeline_ex}
y2 <- y1 %>%
  
  ## Filter
  filter(COUNTRY %in% c("USA", "MEXICO") & year(date) >= 2000) 

## Create a theme
my_theme <-   theme(
  axis.title.y = element_blank(),
  panel.background = element_blank(),
  legend.position="bottom",
  legend.key = element_blank(),
  axis.line.x = element_line()
)

plot0 <- ggplot() +
  geom_timeline(data = y2, aes(
    x = date,
    xmin = as.Date("2005-01-01"), 
    xmax = as.Date("2018-02-22"), 
    colour = TOTAL_DEATHS, 
    y = COUNTRY, 
    size = EQ_PRIMARY
    )
  ) +
  scale_size_continuous(name = "Richter scale value") +
  scale_colour_continuous(name = "# deaths") +
  xlab("DATE") +
  my_theme
plot0
```

## `geom_timeline_label()`
`geom_timeline_label()` provides additional labeling and subsetting capabilities. Users can set a maximum number of earthquakes to plot per country using the `n_max` parameter (the function selects the highest magnitude earthquakes first). The output also provides location labels for many earthquakes. Below is a graph of the same set of earthquakes as the prior example, selecting only the 10 highest magnitude earthquakes.

```{r geom_timeline_label_ex}
plot1 <- ggplot() +
  geom_timeline_label(data = y2, aes(
    x = date, 
    xmin = as.Date("2005-01-01"), 
    label = LOCATION_NAME,
    xmax = as.Date("2018-02-22"), 
    colour = TOTAL_DEATHS, 
    y = COUNTRY, 
    size = EQ_PRIMARY
    ),
    n_max = 10
  ) +
  scale_x_date(expand = c(.15, .15)) +
  scale_size_continuous(name = "Richter scale value") +
  scale_colour_continuous(name = "# deaths") +
  xlab("DATE") +
  my_theme
plot1
```

## `eq_create_label()`

`eq_create_label()` creates a popup label column in the earthquake data for use in `eq_map()` (discussed below). The label displays the earthquake location, magnitude, and death toll--if they are available--with HTML formatting.
```{r eq_create_label_ex}
y3 <- y2 %>%
  mutate(popup_text = eq_create_label(.))
head(y3$popup_text)
```

## `eq_map`

`eq_map` shows earthquakes on an interactive map, with more information the earthquake provided when users click on the plotted point. Users can provide arguments to the `annot_col` parameter to choose a variable that serves as the popup text label. Below is an example map using the same set of earthquakes as the first plot using the `popup_text` variable created in the above example.

```{r eq_map_ex}
y3 %>%
  eq_map(annot_col = "popup_text")
```
